Northern Pikeminnow Abundance in Deadwater Slough, Salmon River, Idaho and Potential Impacts to Local Chinook Salmon Populations

### 2022-02-03 analysis/paper/paper.Rmd
latest edits from TB

Quantifying mortality associated with piscine predation in upstream habitats of the Salmon River is of particular interest for spring/summer-run Chinook Salmon (hereafter Chinook Salmon) in the Upper Salmon major population group (MPG). The Salmon River was historically the most productive tributary for Chinook Salmon in the Columbia River basin [@Nemeth1999], but those populations have become depleted in recent decades. Currently, the Upper Salmon MPG supports eight independent, extant populations, including Salmon River (above Redfish Lake Creek), Valley Creek, Yankee Fork Salmon River, East Fork Salmon River, Salmon River (mainstem below Redfish Lake Creek), Pahsimeroi River, Lemhi River, and North Fork Salmon River [@NOAA2017]. The Lemhi River historically contained the largest Chinook Salmon population within the MPG and is an intensively monitored watershed to measure responses to restoration actions [@Zimmerman2012]. The Upper Salmon MPG are stream-type Chinook Salmon and exhibit two distinct migration tactics; downstream rearing (DSR) and natal reach rearing (NRR) [@Copeland2014a]. The DSR migrants leave their natal area as subyearlings between June and November and typically overwinter in downstream, mainstem habitats until the following spring when they emigrate to the ocean as smolts. Alternatively, NRR migrants remain in their natal areas for approximately one year after emergence until emigration to the ocean as smolts. Diversity of migration tactics provides a mechanism for coping with adverse conditions in freshwater rearing and migration environments and buffers against catastrophic events, thereby increasing population resiliency [@Dodson2013]. Deadwater Slough represents an important habitat for the Upper Salmon MPG as is supports rearing and migration of juveniles from all populations including fall DSR and spring NRR migrants.

@@

Multiple capture methods were employed during the initial fall 2019 survey to reduce selectivity and bias for species and size classes. Methods included raft electrofishing, fyke netting, snorkeling, and angling. Ineffective methods were abandoned in subsequent years. After evaluating all methods, angling was the most effective method for capturing Northern Pikeminnow while also minimizing potential impacts to ESA-listed adult steelhead that were present in the Salmon River during our surveys. Therefore, the following analyses will focus on fish captured by angling, unless otherwise noted.

@@

where $M$, $n$, and $m$ are indexed by each survey (day), $i$. With small numbers of recaptures, it may not be appropriate to calculate the standard error of the Schnabel estimator, but the 95% confidence intervals can be calculated using a Poisson approximation [@Krebs1999]. The Schnabel estimator is essentially the weighted average of a series of Chapman-modified Lincoln-Peterson estimators.

Because captures of new and marked fish were dependent on volunteer anglers, there was concern that sampling was biased to locations preferred by anglers. If marked fish were not recently released or mixed back into those preferred locales, marked fish may have been less susceptible to capture than unmarked fish. To account for this potential bias, we also used an adjusted Schnabel estimator, which assumed that marked and released fish during each survey day were not available for recapture for 48 hours. The adjusted Schnabel estimator, which we refer to as the "delayed mixing" Schnabel estimator, would also account for marked fish being "hook shy" for 48 hours following release.

@@

This approach assumes equal capture probabilities for fall and spring surveys. Fall and spring population estimates were converted into a linear density by dividing by the length of Deadwater Slough (rkm) measured in GIS software and using drone-generated orthography.

@@

The estimate of total juvenile Chinook Salmon biomass consumed was then converted to the total estimated number of fish consumed each period by dividing by the average weight of DSR (`r avg_DSR_g` g) and NRR (`r avg_NRR_g` g) emigrants in the Upper Salmon MPG. The average weights of DSR and NRR emigrants were calculated from fish captured at seven rotary screw traps upstream of Deadwater Slough during the fall and spring periods. Although the primary impact to Chinook Salmon in Deadwater Slough by Northern Pikeminnow is juvenile predation, the common metric to evaluate salmon recovery is often adult returns. We assessed the potential impacts on adult returns by estimating the number of additional adult Chinook Salmon that might be expected to return to Lower Granite Dam if predation in the Deadwater Slough was reduced or eliminated. This calculation was made by multiplying the total estimated juvenile Chinook Salmon consumed by an estimate of Granite-to-Granite smolt-to-adult survival rate (SAR) from @McCann2019 for Upper Salmon River MPG Chinook Salmon.

@@

Mark-recapture abundance estimates for Northern Pikeminnow in Deadwater Slough during fall 2019 ranged from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` and for fall 2020 ranged from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` depending on the estimator used (Table \@ref(tab:abund-tab)). For both surveys, the multiple census estimates were generally larger than the single census estimates and small numbers of recaptures resulted in wide confidence intervals (Figure \@ref(fig:abund-fig)). Because our sampling design most closely matched a multiple census estimator and required a small sample size modification, we deemed the Schnabel estimates most appropriate. Accordingly, the mean Northern Pikeminnow abundance for the two fall sampling events was `r prettyNum(mean(all_est$N[all_est$model == "Schnabel"]), big.mark = ",")` (95% CI: `r paste(prettyNum(round(c(spring_abund$Lci_F, spring_abund$Uci_F)), big.mark = ","), collapse = " - ")`). If we assume that marked and released fish are unavailable for recapture for 48 hours following release to address concerns about delayed mixing into preferred angling locales and/or released fish being "hook shy", the mean fall abundance estimate was `r prettyNum(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` (95% CI: `r prettyNum(mean(all_est$Lci[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` - `r prettyNum(mean(all_est$Uci[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")`).

For spring 2021, we estimated abundance by multiplying the mean fall abundance (standard Schnabel estimates) by the ratio between the spring CPUE and mean fall CPUEs (`r round(total_cpue$catch_per_hr[total_cpue$event_name == "Spring 2021"] / mean(total_cpue$catch_per_hr[total_cpue$event_name %like% "Fall"]), 2)`), resulting in an estimate of `r prettyNum(round(spring_abund$N_S), big.mark = ",")` (95% CI: `r paste(prettyNum(round(c(spring_abund$Lci_S, spring_abund$Uci_S)), big.mark = ","), collapse = " - ")`) Northern Pikeminnow in Deadwater Slough. Those estimates translate to linear densities of `r prettyNum(round(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_rkms), big.mark = ",")` and `r prettyNum(round(spring_abund$N_S / ddw_rkms), big.mark = ",")` Northern Pikeminnow per rkm.

@@

During the fall DSR emigration, we estimate an average-size Northern Pikeminnow (`r round(avg_fall_NPM_TL, 1)` TL mm) to consume `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f60_g"]` g of fish to maintain their body size, assuming 60% of their diet consists of fish prey (Figure \@ref(fig:bio-p)). Fall consumption ranged from `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f30_g"]` g for a diet of 30% fish to `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f90_g"]` g for a diet of 90% fish. During the spring NRR emigration, we estimated an average-size Northern Pikeminnow (`r round(avg_spring_NPM_TL, 1)` TL mm) to consume `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f60_g"]` g of fish, assuming 60% of their diet is fish prey (Figure \@ref(fig:bio-p)). Spring consumption ranged from `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f30_g"]` for a diet of 30% fish to `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f90_g"]` for a diet of 90% fish. Given estimated mean Northern Pikeminnow population sizes of `r prettyNum(mean(all_est$N[all_est$model == "Schnabel"]), big.mark = ",")` from fall estimates and `r prettyNum(round(spring_abund$N_S), big.mark = ",")` for spring 2021, and assuming 60% of their diet is fish, we estimate that Northern Pikeminnow population can consume `r round(tot_g_DSR, 1)` metric tons of fish during the fall DSR emigration and an additional `r round(tot_g_NRR, 1)` metric tons during the NRR emigration in Deadwater Slough.

@@

Using the total fish consumed by the Northern Pikeminnow population (summing fall and spring surveys), we produced a sensitivity analysis for three Northern Pikeminnow diet scenarios where juvenile Chinook Salmon make up 30%, 50%, and 65% of total fish prey consumed (Figure \@ref(fig:impact-p)). Assuming that 50% of the total fish consumed are juvenile Chinook salmon, we estimate that `r prettyNum(round(total_impacts$juv_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` (95% CI: `r prettyNum(round(total_impacts$juv_Lci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` - `r prettyNum(round(total_impacts$juv_Uci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")`) juvenile Chinook Salmon will be consumed. We then used our estimate of total juvenile Chinook Salmon consumed to estimate an "adult equivalents" using the median Granite-to-Granite SAR of `r round(sar$medsar, 5)` (sd = `r format(round(sar$sdsar, 5), scientific = F)`) from @McCann2019. Assuming predation was non-existent in Deadwater Slough, the juvenile consumption would be equivalent to `r prettyNum(round(total_impacts$adult_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` (95% CI: `r prettyNum(round(total_impacts$adult_Lci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` - `r prettyNum(round(total_impacts$adult_Uci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")`) adults.

@@

All mark-recapture abundance estimators suggest a large population of Northern Pikeminnow occupy Deadwater Slough. The mean population size of Northern Pikeminnow was estimated to be `r prettyNum(mean(all_est$N[all_est$model == "Schnabel"]), big.mark = ",")` during the fall DSR emigration and `r prettyNum(round(spring_abund$N_S), big.mark = ",")` during the spring NRR emigration. Those estimates translate to linear densities (fish/rkm) that are `r round((mean(all_est$N[all_est$model == "Schnabel"]) / ddw_rkms) / john_day_dens)` and `r round((spring_abund$N_S / ddw_rkms) / john_day_dens)` times greater than densities reported by @Beamesderfer1991 for the John Day Reservoir and approximately `r round((mean(all_est$N[all_est$model == "Schnabel"]) / ddw_rkms) / low_columbia_fish_km)` and `r round((spring_abund$N_S / ddw_rkms) / low_columbia_fish_km)` times greater than projections for the lower Columbia River [@Beamesderfer1996]. Our areal densities (fish/100 $m^2$) for Deadwater Slough are `r round((mean(all_est$N[all_est$model == "Schnabel"]) / ddw_m2) * 100, 1)` and `r round((spring_abund$N_S / ddw_m2) * 100, 1)` for fall and spring, respectively. @Beamesderfer1991 acknowledge that their methods were unsuitable for sampling offshore, noting that water velocity, depth, irregular bottom contours, and barge traffic made sampling offshore ineffective; therefore it is possible that their estimates may have been low. Conversely, at Deadwater Slough the maximum depth was approximately 6 m with a relatively homogenous bottom contour, making angling an effective method throughout the entire reach. Given the size (volume) of the Salmon River at the slough relative to the Columbia River, the estimated Northern Pikeminnow densities for Deadwater Slough are remarkable. Further, our estimate densities of Northern Pikeminnow suggest that slow water reaches outside of the reservoir complexes on the Snake and Columbia rivers may support exceptionally high predator densities, consistent with findings of @Harnish2014, @Gray2001, and @Zimmerman1999. Given the dearth of current data available in the Columbia River basin, it is unclear how Northern Pikeminnow abundance estimates in this study compare to elsewhere in the Columbia River Basin; contemporary information on piscine predator abundances may be needed [@Widener2021].

Several assumptions in our estimators may have influenced the magnitude of our abundance results. First, all models assumed a closed population, however, it is possible that immigration and emigration occurred. Nevertheless, our estimators still provide unbiased estimates of abundance assuming the immigration and emigration rates are equal between marked and unmarked fish. Given the size of the sampling area, and the short duration of our surveys, that assumption is likely met. We also caught one individual that was confirmed tagged in Deadwater Slough in a previous year suggesting that at least some individuals remain in the reach or at least migrate to Deadwater Slough multiple times, annually. The assumption of a closed population can also affect results if mortality occurs for some marked fish released back to the population. During field processing, no mortalities or injury post release was observed; however, the potential for mortality amongst marked and released fish cannot be discounted. Next, our estimates assumed that capture and recapture events are random samples of the population. Angling methods may have a size selection bias, thereby limiting our abundance estimates to a size range susceptible to angling, resulting in a conservative estimate of the total population. We also question the validity of our assumption of equal catchability of individual fish between sampling events. During our study, we recaptured two individual fish during multiple days within a survey indicating that marked fish were still susceptible to angling. We are also comfortable with the assumptions of no loss of marks or no mistaken or missed marks given that marks were physical (hole punches in fins) and that each fish was examined on a measuring board to collect lengths. However, we suspect that marked and released fish may not have mixed thoroughly back into the population prior to recapture events. Although we attempted to release marked fish evenly throughout Deadwater Slough, we anecdotally observed anglers congregating in particular areas of the slough to socialize or exploit the "good" fishing locales. Violations of the assumption of equal catchability between marked and unmarked fish may lead to an overestimate of abundance of Northern Pikeminnow in our study if marked fish have lower catchability. This bias will be proportional to the difference in catchability of marked and unmarked fish. As an example, if marked fish are only 70% as likely to be caught as unmarked fish (due to behavior or spatial distribution) the true abundance is closer to 70% of our abundance estimate.

@@

Our approach assumed that all available prey consumed by Northern Pikeminnow have energy densities equal to juvenile Chinook Salmon. Generalizing energy densities using a single species is a common approach [@Petersen1999], however, we recognize that other salmonids including juvenile steelhead and Sockeye Salmon are likely also available prey in Deadwater Slough, especially during the spring migration. Additionally, hatchery smolt releases (Chinook Salmon, steelhead, and Sockeye Salmon) are also prevalent in the mainstem Salmon River during the spring migration. To address the prevalence of other fish in the Northern Pikeminnow's diet, we considered scenarios where DSR and NRR Chinook Salmon represent only 30%, 50%, and 65% of total fish prey consumed. These values may be considered conservative during the emigration periods when Chinook Salmon are abundant and the Northern Pikeminnow diet shifts almost entirely to piscivory, presumably on juvenile Chinook Salmon [@Shively1996; @Poe1991]. This is likely the case during the fall when DSR Chinook Salmon are the dominant prey species available in Deadwater Slough as 1) no (or at least minimal) hatchery releases are present in the river and 2) fall emigrations of steelhead and Sockeye Salmon are less prominent that in Chinook Salmon. During spring when natural-origin steelhead and Sockeye Salmon are also actively emigrating from the Upper Salmon River and hatchery-origin releases of all three species (Chinook Salmon, steelhead, Sockeye Salmon) are present in the river, <50% Chinook Salmon in their diet may be more likely. Future work to quantify and identify the juvenile salmonid in Deadwater Slough during their seasonal migration would be useful to validate the diet composition assumptions used in our model.

@@

Spawner abundance is perhaps the most important metric considered in determining a population's viability. We estimated that consumption of juvenile Chinook Salmon by Northern Pikeminnow in the Deadwater Slough potentially reduces natural-origin returns to upriver populations by `r round(total_impacts$adult_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50])` (95% CI: `r round(total_impacts$adult_Lci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50])` - `r round(total_impacts$adult_Uci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50])`) adults, annually. This is `r round((total_impacts$adult_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50] / 558) * 100)`% of the total mean adult escapement estimated<!--of 558--> for the Upper Salmon MPG in 2017 - 2019 [@Kinzer2020]. Even under the most conservative scenario where 30% of the Northern Pikeminnow diet is fish and 30% of that is Chinook Salmon we estimate the "adult equivalents" of the juveniles consumed to be `r round(total_impacts$adult_N[total_impacts$perc_diet == 30 & total_impacts$p_chnk == 0.30])` (95% CI: `r round(total_impacts$adult_Lci[total_impacts$perc_diet == 30 & total_impacts$p_chnk == 0.30])` - `r round(total_impacts$adult_Uci[total_impacts$perc_diet == 30 & total_impacts$p_chnk == 0.30])`) which is `r round((total_impacts$adult_N[total_impacts$perc_diet == 30 & total_impacts$p_chnk == 0.30] / 558) * 100)`% of recent natural-origin adult escapements. In addition to juvenile Chinook Salmon, juvenile steelhead and Sockeye Salmon, including hatchery releases, are also likely prey items for Northern Pikeminnow in Deadwater Slough. Accordingly, we surmise that adult returns of steelhead and Sockeye Salmon are also affected by predation, including hatchery populations, which provide recreational fishing opportunities. Consequently, reducing predation mortality at Deadwater Slough could potentially benefit multiple upriver natural and hatchery populations, including other ESA-listed species. Because the Deadwater Slough is part of the migratory pathway for multiple species and populations of emigrating salmonids, the positive impact could be greater than individual tributary rehabilitation actions which typically benefit a single population.

@@

```{r cpue-fig, fig.cap = "The number of Northern Pikeminnow caught per angler effort for each sample effort.", eval = F}
total_cpue %>%
  ggplot(aes(x = event_name,
             y = catch_per_hr)) +
  geom_col(fill = "gray30") +
  labs(y = "Catch Per Angler Hour") +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_text(size = 11,
                                   color = "black"),
        axis.title.y = element_text(vjust = 3),
        axis.text.y = element_text(size = 11,
                                   color = "black"))

```{r abund-fig, fig.cap = "Estimates of abundance of Northern Pikeminnow in Deadwater Slough using different mark-recapture estimators and for the fall sampling events."}

>>>>

Quantifying mortality associated with piscine predation in upstream habitats of the Salmon River is of particular interest for spring/summer-run Chinook Salmon (hereafter Chinook Salmon) in the Upper Salmon major population group (MPG). The Salmon River was historically the most productive tributary for Chinook Salmon in the Columbia River basin [@Nemeth1999], but those populations have become depleted in recent decades. Currently, the Upper Salmon MPG supports eight independent, extant populations; this includes Salmon River (above Redfish Lake Creek), Valley Creek, Yankee Fork Salmon River, East Fork Salmon River, Salmon River (mainstem below Redfish Lake Creek), Pahsimeroi River, Lemhi River, and North Fork Salmon River populations [@NOAA2017]. The Lemhi River historically contained the largest Chinook Salmon population within the MPG and is therefore of particular interest for restoration actions [@Zimmerman2012]. The Upper Salmon MPG are stream-type Chinook Salmon and exhibit two distinct migration tactics; downstream rearing (DSR) and natal reach rearing (NRR) [@Copeland2014a]. The DSR migrants leave their natal area as subyearlings between June and November and typically overwinter in downstream, mainstem habitats until the following spring when they emigrate to the ocean as smolts. Alternatively, NRR migrants remain in their natal areas for approximately one year after emergence until emigration to the ocean as smolts. Diversity of migration tactics provides a mechanism for coping with adverse conditions in freshwater rearing and migration environments and buffers against catastrophic events, thereby increasing population resiliency [@Dodson2013]. Deadwater Slough represents an important habitat for the Upper Salmon MPG as is supports rearing and migration of juveniles from all populations including fall DSR and spring NRR migrants.

@@

Multiple capture methods were employed during the initial fall 2019 survey to reduce selectivity and bias for species and size classes. Methods included raft electrofishing, fyke netting, snorkeling, and angling. After evaluating all methods, angling was the most effective method for capturing Northern Pikeminnow while also minimizing potential impacts to ESA-listed adult steelhead that were present in the Salmon River during our surveys. Ineffective methods were abandoned in subsequent years and the following analyses will focus on fish captured by angling, unless otherwise noted.

@@

where $M$, $n$, and $m$ are indexed by each survey (day), $i$. With small numbers of recaptures, it may not be appropriate to calculate the standard error of the Schnabel estimator, but the 95% confidence intervals can be calculated using a Poisson approximation [@Krebs1999].<!--The Schnabel estimator is essentially the weighted average of a series of Chapman-modified Lincoln-Peterson estimators.-->

Because captures of new and marked fish were dependent on volunteer anglers, there was concern that sampling was biased to locations preferred by anglers. If marked fish were not recently released or mixed back into those preferred locales, marked fish may have been less susceptible to capture than unmarked fish. To account for this potential bias, we also used an adjusted Schnabel estimator, which assumed that marked and released fish during each survey day were not available for recapture for 48 hours. The adjusted Schnabel estimator, referred to as the "delayed mixing" Schnabel estimator, is given by:

@@

This approach assumes equal capture probabilities for fall and spring surveys. Fall and spring population estimates were converted into linear and areal densities by dividing by the length and area, respectively, of Deadwater Slough measured in GIS software and using drone-generated orthography.

@@

The estimate of total juvenile Chinook Salmon biomass consumed was then converted to the total estimated number of fish consumed each period by dividing by the average weight of DSR (`r avg_DSR_g` g) and NRR (`r avg_NRR_g` g) emigrants in the Upper Salmon MPG. The average weights of DSR and NRR emigrants were calculated from fish captured at seven rotary screw traps upstream of Deadwater Slough during the fall and spring periods. Although the primary impact to Chinook Salmon in Deadwater Slough by Northern Pikeminnow is juvenile predation, the common metric to evaluate salmon recovery is often adult returns. We assessed the potential impacts on adult returns by estimating the number of additional adult Chinook Salmon expected to return to Lower Granite Dam if predation was reduced or eliminated in the Deadwater Slough. This calculation was made by multiplying the total estimated juvenile Chinook Salmon consumed by the median Granite-to-Granite SAR of 0.00614 (sd = 0.00051) from @McCann2019 for Upper Salmon River MPG Chinook Salmon.

@@

Mark-recapture abundance estimates for Northern Pikeminnow in Deadwater Slough during fall 2019 ranged from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` and for fall 2020 ranged from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` depending on the estimator used (Table \@ref(tab:abund-tab)). For both surveys, the multiple census estimates were generally larger than the single census estimates and small numbers of recaptures resulted in wide confidence intervals (Figure \@ref(fig:abund-fig)). Because our sampling design most closely matched a multiple census estimator and required a small sample size modification, we deemed the Schnabel estimates most appropriate. Accordingly, the mean Northern Pikeminnow abundance for the two fall sampling events was `r prettyNum(mean(all_est$N[all_est$model == "Schnabel"]), big.mark = ",")` (95% CI: `r paste(prettyNum(round(c(spring_abund$Lci_F, spring_abund$Uci_F)), big.mark = ","), collapse = " - ")`). Using the delayed-mixing Schnabel estimator, the mean fall abundance estimate was `r prettyNum(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` (95% CI: `r prettyNum(mean(all_est$Lci[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` - `r prettyNum(mean(all_est$Uci[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")`). For spring 2021, we estimated `r prettyNum(round(spring_abund$N_S), big.mark = ",")` (95% CI: `r paste(prettyNum(round(c(spring_abund$Lci_S, spring_abund$Uci_S)), big.mark = ","), collapse = " - ")`) Northern Pikeminnow in Deadwater Slough. Those estimates translate to linear densities of `r prettyNum(round(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_rkms), big.mark = ",")` and `r prettyNum(round(spring_abund$N_S / ddw_rkms), big.mark = ",")` Northern Pikeminnow per rkm and areal densities (fish/100 m$^2$) of `r round((mean(all_est$N[all_est$model == "Schnabel"]) / ddw_m2) * 100, 1)` and `r round((spring_abund$N_S / ddw_m2) * 100, 1)` for fall and spring, respectively.

@@

During the fall DSR emigration, we estimate an average-size Northern Pikeminnow (`r round(avg_fall_NPM_TL, 1)` TL mm) to consume `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f60_g"]` g of fish to maintain their body size, assuming 60% of their diet consists of fish prey (Figure \@ref(fig:bio-p)). Fall consumption ranged from `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f30_g"]` g for a diet of 30% fish to `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f90_g"]` g for a diet of 90% fish. During the spring NRR emigration, we estimated an average-size Northern Pikeminnow (`r round(avg_spring_NPM_TL, 1)` TL mm) to consume `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f60_g"]` g of fish, assuming 60% of their diet is fish prey (Figure \@ref(fig:bio-p)). Spring consumption ranged from `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f30_g"]` for a diet of 30% fish to `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f90_g"]` for a diet of 90% fish. Given estimated mean Northern Pikeminnow population sizes of `r prettyNum(mean(all_est$N[all_est$model == "Schnabel"]), big.mark = ",")` from fall estimates and `r prettyNum(round(spring_abund$N_S), big.mark = ",")` for spring 2021, and assuming 60% of their diet is fish, we estimate that Northern Pikeminnow population can consume `r round(tot_g_DSR, 1)` metric tons of fish during the fall DSR emigration and an additional `r round(tot_g_NRR, 1)` metric tons during the spring NRR emigration in Deadwater Slough.

@@

Using the total fish consumed by the Northern Pikeminnow population (summing fall and spring surveys), we produced a sensitivity analysis for three Northern Pikeminnow diet scenarios where juvenile Chinook Salmon make up 30%, 50%, and 65% of total fish prey consumed (Figure \@ref(fig:juv-con-p)). Assuming that 50% of the total fish consumed are juvenile Chinook salmon, we estimate that `r prettyNum(round(total_impacts$juv_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` (95% CI: `r prettyNum(round(total_impacts$juv_Lci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` - `r prettyNum(round(total_impacts$juv_Uci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")`) juvenile Chinook Salmon will be consumed. Using the median Granite-to-Granite SAR, we estimated the juvenile consumption would is equivalent to `r prettyNum(round(total_impacts$adult_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` (95% CI: `r prettyNum(round(total_impacts$adult_Lci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` - `r prettyNum(round(total_impacts$adult_Uci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")`) adults (Figure \@ref(fig:adult-imp-p)).

@@

All mark-recapture abundance estimators suggest a large population of Northern Pikeminnow occupy Deadwater Slough. The mean population size of Northern Pikeminnow was estimated to be `r prettyNum(mean(all_est$N[all_est$model == "Schnabel"]), big.mark = ",")` during the fall DSR emigration and `r prettyNum(round(spring_abund$N_S), big.mark = ",")` during the spring NRR emigration. Those estimates translate to linear densities (fish/rkm) that are `r round((mean(all_est$N[all_est$model == "Schnabel"]) / ddw_rkms) / john_day_dens)` and `r round((spring_abund$N_S / ddw_rkms) / john_day_dens)` times greater than densities reported by @Beamesderfer1991 for the John Day Reservoir and approximately `r round((mean(all_est$N[all_est$model == "Schnabel"]) / ddw_rkms) / low_columbia_fish_km)` and `r round((spring_abund$N_S / ddw_rkms) / low_columbia_fish_km)` times greater than projections for the lower Columbia River [@Beamesderfer1996]. @Beamesderfer1991 acknowledge that their methods were unsuitable for sampling offshore, noting that water velocity, depth, irregular bottom contours, and barge traffic made sampling offshore ineffective; therefore it is possible that their estimates may have been low. Conversely, at Deadwater Slough the maximum depth was approximately 6 m with a relatively homogenous bottom contour, making angling an effective method throughout the entire reach. Given the size (volume) of the Salmon River at the slough relative to the Columbia River, the estimated Northern Pikeminnow densities for Deadwater Slough are remarkable. Further, our estimate densities of Northern Pikeminnow suggest that slow water reaches outside of the reservoir complexes on the Snake and Columbia rivers may support exceptionally high predator densities, consistent with findings of @Harnish2014, @Gray2001, and @Zimmerman1999. Given the dearth of current data available in the Columbia River basin, it is unclear how Northern Pikeminnow abundance estimates in this study compare to elsewhere in the Columbia River Basin; contemporary information on piscine predator abundances may be needed [@Widener2021].

Several assumptions in our estimators may have influenced the magnitude of our abundance results. First, all models assumed a closed population, however, it is possible that immigration and emigration occurred. Nevertheless, our estimators still provide unbiased estimates of abundance assuming the immigration and emigration rates are equal between marked and unmarked fish. Given the size of the sampling area, and the short duration of our surveys, that assumption is likely met. <!--We also caught one individual that was confirmed tagged in Deadwater Slough in a previous year suggesting that at least some individuals remain in the reach or at least migrate to Deadwater Slough multiple times, annually.-->The assumption of a closed population can also affect results if mortality occurs for some marked fish released back to the population. During field processing, no mortalities or injury post release was observed; however, the potential for mortality amongst marked and released fish cannot be discounted. Next, our estimates assumed that capture and recapture events are random samples of the population. Angling methods may have a size selection bias, thereby limiting our abundance estimates to a size range susceptible to angling, resulting in a conservative estimate of the total population. We also question the validity of our assumption of equal catchability of individual fish between sampling events. During our study, we recaptured two individual fish during multiple days within a survey indicating that marked fish were still susceptible to angling.<!--We are also comfortable with the assumptions of no loss of marks or no mistaken or missed marks given that marks were physical (hole punches in fins) and that each fish was examined on a measuring board to collect lengths.--> However, we suspect that marked and released fish may not have mixed thoroughly back into the population prior to recapture events. Although we attempted to release marked fish evenly throughout Deadwater Slough, we anecdotally observed anglers congregating in particular areas of the slough to socialize or exploit the "good" fishing locales. Violations of the assumption of equal catchability between marked and unmarked fish may lead to an overestimate of abundance of Northern Pikeminnow in our study if marked fish have lower catchability. This bias will be proportional to the difference in catchability of marked and unmarked fish. For example, if marked fish were only 70% as likely to be caught as unmarked fish, the true abundance is closer to 70% of our abundance estimate.

@@

Our approach assumed that all available prey consumed by Northern Pikeminnow have energy densities equal to juvenile Chinook Salmon. Generalizing energy densities using a single species is a common approach [@Petersen1999], however, we recognize that other salmonids including juvenile steelhead and Sockeye Salmon are likely also available prey in Deadwater Slough, especially during the spring migration. Additionally, hatchery smolt releases (Chinook Salmon, steelhead, and Sockeye Salmon) are also prevalent in the mainstem Salmon River during the spring migration. To address the prevalence of other fish in the Northern Pikeminnow's diet, we considered scenarios where DSR and NRR Chinook Salmon represent only 30%, 50%, and 65% of total fish prey consumed. These values may be considered conservative during the emigration periods when Chinook Salmon are abundant and the Northern Pikeminnow diet shifts almost entirely to piscivory, presumably on juvenile Chinook Salmon [@Shively1996; @Poe1991]. This is likely the case during the fall when DSR Chinook Salmon are the dominant prey species available in Deadwater Slough as 1) no (or at least minimal) hatchery releases are present in the river and 2) fall emigrations of steelhead and Sockeye Salmon are less prominent that in Chinook Salmon. During spring when natural-origin steelhead and Sockeye Salmon are also actively emigrating from the Upper Salmon River and hatchery-origin releases of all three species (Chinook Salmon, steelhead, Sockeye Salmon) are present in the river, <50% Chinook Salmon in the Pikeminnow's diet may be more likely. Future work to quantify and identify the juvenile salmonid in Deadwater Slough during their seasonal migration would be useful to validate the diet composition assumptions used in our model.

@@

Spawner abundance is perhaps the most important metric considered in determining a population's viability. We estimated that consumption of juvenile Chinook Salmon by Northern Pikeminnow in the Deadwater Slough potentially reduces natural-origin returns to upriver populations by `r round(total_impacts$adult_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50])` adults, annually, which is `r round((total_impacts$adult_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50] / 558) * 100)`% of the total mean adult escapement estimated<!--of 558--> for the Upper Salmon MPG in 2017 - 2019 [@Kinzer2020]. Even under the most conservative scenario where 30% of the Northern Pikeminnow diet is fish and 30% of that is Chinook Salmon we estimate the "adult equivalents" of the juveniles consumed to be `r round(total_impacts$adult_N[total_impacts$perc_diet == 30 & total_impacts$p_chnk == 0.30])`, which is `r round((total_impacts$adult_N[total_impacts$perc_diet == 30 & total_impacts$p_chnk == 0.30] / 558) * 100)`% of recent natural-origin adult escapements. In addition to juvenile Chinook Salmon, juvenile steelhead and Sockeye Salmon, including hatchery releases, are also likely prey items for Northern Pikeminnow in Deadwater Slough. Accordingly, we surmise that adult returns of steelhead and Sockeye Salmon are also affected by predation, including hatchery populations, which provide recreational fishing opportunities. Consequently, reducing predation mortality at Deadwater Slough could potentially benefit multiple upriver natural and hatchery populations, including other ESA-listed species. Because the Deadwater Slough is part of the migratory pathway for multiple species and populations of emigrating salmonids, the positive impact could be greater than individual tributary rehabilitation actions which typically benefit a single population.

@@

```{r abund-fig, fig.cap = "Estimates of abundance of Northern Pikeminnow in Deadwater Slough using different mark-recapture estimators and for the fall sampling events. Error bars indicate 95% confidence intervals."}
